First paper - Draft

### 2017-07-03 paper/paper.Rnw
Playing around with latex tables

<<working directory, echo = F>>=
options(scipen = 213141)
opts_chunk$set(echo = F,
               message = F,
               warning = F,
               include = F,
               cache.lazy = F,
               results = 'asis')
@

<<default_conf>>=
  library(knitr)
  library(arm)
  library(saves)
  library(haven)
  library(PISA2000lite)
  library(PISA2003lite)
  library(PISA2006lite)
  library(PISA2009lite)
  library(PISA2012lite)
  library(intsvy)
  library(cimentadaj) # devtools::install_github("cimentadaj/cimentadaj")
  library(countrycode) # For region variable
  library(car)
  library(SAScii)
  library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
  library(lme4)
  library(modelr)
  library(tidyverse)
  library(ggrepel)

@@

<<loading_data-recoding>>=
pisa_all <- read_rds("./data/pisa_listcol.Rdata")
pisa_all2 <- pisa_all

@@

<<escs_trend, cache = TRUE>>=
  # Rescaled trend ESCS data to merge.
  # This only has data for seq(2000, 2012, 3) because
  # PISA 2015 has the ESCS trend variable.
  dir <- tempdir()
  file_name <- "escs_trend.zip"
  download.file("http://vs-web-fs-1.oecd.org/pisa/trend_escs_SPSS.zip",
                destfile = file.path(dir, file_name))
  unzip(file.path(dir, file_name), exdir = dir)
  escs_trend <- map(file.path(dir, list.files(dir, pattern = ".sav")), haven::read_spss)
  file.remove(file.path(dir, list.files(dir)))

@@

<<merge_escs_pisa, cache = TRUE>>=
   # Next we'll merge the ESCS data with the PISA data. As explained above, the 6th data (PISA
  # 2015) doesn't need to be merged so I exclude it with this vector
  exclude <- -6

@@

<<modeling, cache = TRUE>>=
adapted_year_data <-
    map(pisa_all2$value, ~ {
      if (unique(.x$wave) == "pisa2000") {
        # pisa2000 has a different coding so here I recode 6 to 7 so that in all waves the top edu
        # is 7 and the bottom is 1
        .x <-
          mutate(.x, new_hisced = as.character(dplyr::recode(as.numeric(high_edu_broad), `6` = 7)))
      } else {
        .x <-
          mutate(.x, new_hisced = as.character(high_edu_broad))
      }
      .x
})

@@

For Denmark, for example, we learned a great deal about how to improve social mobility by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility for less advantaged children \citep{bjorklund2009, jaeger2007}. But the important finding came when we discovered the main reason behind their social escalator: its' educational system. For example, research by \citet{esping2012} and \citet{bauchmuller2015}, show that the Danish early education system has important and longstanding impacts in helping children reach their fullest potential, specially disadvantaged children. Moreover, it tracks students into different curricula very late compared to other Europen countries Europe (age 16). These two traits make the Danish system uniquely effective. The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable. In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.

@@

\citet{reardon2011} finds that the increase in the gaps has occured predominantely from the 1970's until the 2000's. In fact, the hard numbers suggest that the gap widened by \~40-50\%. The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of \~75\%. Given that the studies before 1970 are less reliablle in terms of comparability and sampling design, the author computes all results for both before/after 1970. To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations in the year 2010. Using longitudinal data, \citet{bradbury2015} find similar results. Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25. Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of \~1.25-1.50 standard deviations. One important drawback of these studies is that they don't present the uncertainty of this estimates. Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy. It could be that the gap is at \~1.25 standard deviations, varying up to 1.75 and down to 0.80. Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.

@@

The American social landscape is very distinctive relative to Europe. The reasons the authors find a reversal in the trend could be numerous and should be studied very close. For example, the American educational system lacks any formal curricular differentiation \emph{\à la} European style. It is true nonetheless that \citet{reardon2011} did not cover from the period 2000 onwards, which is specifically the trend they studied so we have no other comparison. \citet{reardonportilla} suggests that the reversal is likely due to the high increase of preschool enrollment. They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially. This explanation still needs further confirmation.

Motivated by these recent results, other authors have taken this analysis to an international context in order to discover between-country trends. The work of \citet{bradbury2015} employs a unique comparative analysis of the achievement gap between Australia, United Kingdom, United States and Canada. Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is not only the first study that uses panel data to study achievement gaps, but to also do it between countries}. The core finding behind their study is that the American achievement gap is much wider than in any other comparison country, specially Australia and Canada. They find that once the achievement gap is present in early school entry, it doesn't seem to narrow or neither widen very much over the life course. In fact, they calculate that the schooling environment once they enter school can only explain about \~30-40\% of the high school SES gap. This suggests that once the achievement gap widens before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous RCT's show that high quality schooling can indeed help ease the gap, in some instances even close it \citep{campbell2002}.}. One exception is the United Kingdom, which they found to help close the gap in early primary years. This can likely be due to the comprehensive schooling and also the public support by the welfare state in dimensions like health and welfare support.

@@

More formally, We're interested in studying
a) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
b) Which countries are experiencing changes in the achievement gap;
c) the rate at which each country-gap is widening/narrowing and;
d) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.

@@

a) We want to test the notion that better performing countries have a smaller degree of inequality than other countries. \citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious, however. For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.

b) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data. This shows that it is important to study the changes in the achievement gap over time. This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.

c) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year available. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}

d) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving. If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.

@@

This new standardized variable has a mean of zero. Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves because the test scores have now the same metric caross time. However, if the variance of academic achievement changes over time, then standardizing the overall score at each country-wave pair actually makes the transformation biased. That is, by standardizing we're forcing the standard deviation of test scores to be zero across all waves. But if the true deviations of the median academic achievement changes over time, then the estimated trend in the SES gaps will be underestimated, or vice versa.

% Show graph of variance over time
% I'm showing standard deviations, is it better to show variance?
<<variance_pisa, cache = TRUE, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=

@@

var_years %>%
  setNames(seq(2000, 2015, 3)) %>%
  enframer("name") %>%
  ggplot(aes(name, value)) + geom_col()
@

@@

Another concern is that if the gaps at different waves are measured with tests that have different amounts of measurement error, then the amount of bias will not be the same in each measure of the gap. This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citep{reardon2011}. PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but we still have to adjust for this imprecision.

@@

Finally, we calculate the SES achievement gap for each country by extracting the fitted \(\alpha\) and \(\beta\) for each country \(j\) and calculating the difference of the predicted score of High and Low SES.

@@

<<escsdummy_charachteristics, cache = TRUE, eval = F>>=
.x <- adapted_year_data[[1]]
country <- filter(.x, country == "United States")
probs <- c(0.1, 0.9)

escs_dummy_creator <- function(df, probs) {

@@

escs_data[[1]] %>%
  select(country, escs_dummy, high_edu_broad, ISEI) %>%
  filter(country %in% countries) %>%
  count(country, escs_dummy, high_edu_broad) %>%
  group_by(country, escs_dummy) %>%
  mutate(total_n = sum(n),
         perc = n / total_n * 100) %>%
  select(country, escs_dummy, high_edu_broad, perc) %>%
  map_if(is_double, round, 2) %>%
  as_tibble() %>%
  filter(!is.na(escs_dummy)) %>%
  spread(high_edu_broad, perc)

escs_data[[1]] %>%
  select(country, escs_dummy, high_edu_broad, ISEI) %>%
  filter(country %in% countries) %>%
  count(country, escs_dummy, isei_group = cut_number(ISEI, 4)) %>%
  group_by(country, isei_group) %>%
  mutate(total_n = sum(n),
         perc = n / total_n * 100) %>%
  select(country, escs_dummy, isei_group, perc) %>%
  map_if(is_double, round, 2) %>%
  as_tibble() %>%
  filter(!is.na(escs_dummy)) %>%
  spread(isei_group, perc) %>%
  left_join(m, by = c("country", "escs_dummy"))

% Figure out how to ouput data frames as latex tables.
<<sample_size, cache = TRUE>>=
# Get sample counts for each dummy
sample_size_calc <- function(df, probs, selected = F, cnts = NULL) {

@@

sample_tables_topbottom <- sample_size_calc(adapted_year_data, c(.1, .9), selected = TRUE, countries)
sample_tables_topmid <- sample_size_calc(adapted_year_data, c(.5, .9), selected = TRUE, countries)
sample_tables_midbottom <- sample_size_calc(adapted_year_data, c(.1, .5), selected = TRUE, countries)
@

<<merge_math_read, cache = TRUE>>=

@@

% How to output data frames are latex tables?
<<rate_change, cache = TRUE>>=

@@

full_data %>%
  ggplot(aes(low_increase, high_increase), alpha = 0.2) +
  geom_rect(data = rect_data, aes(xmin = xst,
                                  xmax = xen,
                                  ymin = yst,
                                  ymax = yen),
            fill = rect_data$colour,
            alpha = 0.2,
            inherit.aes = FALSE) +
  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "grey", size = 1,
            linetype = "longdash") +
  geom_point(alpha = 0.2) +
  geom_point(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.7) +
  geom_line(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.4,
            stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "red", size = 1) +
  geom_text_repel(data = filter(full_data, continent == "my_cnt"),
                  aes(label = country), box.padding = unit(2.7, "lines")) +
  geom_vline(xintercept = 0, alpha = 0.5) +
  geom_hline(yintercept = 0, alpha = 0.5) +
  xlim(lims$xlim) +
  ylim(lims$ylim) +
  coord_cartesian(expand = FALSE) +
  annotate(geom = "text", x = 0.15, y = -0.2,
           label = "Low SES is catching up \n faster than High SES",
           fontface = 2, size = 3) +
  annotate(geom = "text", x = -0.05, y = 0.20,
           label = "High SES is increasing  \n faster than Low SES",
           fontface = 2, size = 3) +
  labs(x = "Average increase of low SES in SD", y = "Average increase of high SES in SD") +
  theme_minimal()
@

In a more convenient way, we can see that there is a lot of variance in the rate at which top/bottom groups are changing. We see that there are more countries where the inequality is increasing (red panel) but there's also a fair share where Low SES is catching up faster, such as United States and Germany. We see a negative relationship where the faster the low SES catch up the slower the High SES decrease. Quite an interesting pattern because we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel). On the other hand, we do see more ocuntries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.

% How to output data frames are latex tables?
<<perc_increase_tables, cache = TRUE>>=
# Show the rates at which is increasing/decreasing
perc_increase_fun <- function(df) {

>>>>

<<working directory, echo = F>>=
options(scipen = 213141)

opts_chunk$set(echo = FALSE,
               message = FALSE,
               cache = TRUE,
               warning = FALSE,
               include = FALSE,
               cache.lazy = FALSE,
               results = 'asis')
@

<<default_conf>>=
  library(knitr)
  library(arm)
  library(saves)
  library(haven)
  library(PISA2000lite)
  library(PISA2003lite)
  library(PISA2006lite)
  library(PISA2009lite)
  library(PISA2012lite)
  library(intsvy)
  library(cimentadaj) # devtools::install_github("cimentadaj/cimentadaj")
  library(countrycode) # For region variable
  library(car)
  library(SAScii)
  library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
  library(lme4)
  library(modelr)
  library(tidyverse)
  library(ggrepel)
  library(xtable)

@@

<<loading_data-recoding>>=
pisa_all <- read_rds("./data/pisa_listcol.Rdata")

pisa_all2 <- pisa_all

@@

<<escs_trend>>=
  # Rescaled trend ESCS data to merge.
  # This only has data for seq(2000, 2012, 3) because
  # PISA 2015 has the ESCS trend variable.
  dir <- tempdir()
  file_name <- "escs_trend.zip"
  download.file("http://vs-web-fs-1.oecd.org/pisa/trend_escs_SPSS.zip",
                destfile = file.path(dir, file_name))
  unzip(file.path(dir, file_name), exdir = dir)
  escs_trend <- map(file.path(dir, list.files(dir, pattern = ".sav")), haven::read_spss)
  file.remove(file.path(dir, list.files(dir)))

@@

<<merge_escs_pisa>>=
   # Next we'll merge the ESCS data with the PISA data. As explained above, the 6th data (PISA
  # 2015) doesn't need to be merged so I exclude it with this vector
  exclude <- -6

@@

<<modeling>>=
adapted_year_data <-
    map(pisa_all2$value, ~ {
      if (unique(.x$wave) == "pisa2000") {
        # pisa2000 has a different coding so here I recode 6 to 7 so that in all waves the top edu
        # is 7 and the bottom is 1
        .x <-
          mutate(.x, new_hisced = as.character(dplyr::recode(as.numeric(high_edu_broad), `6` = 7)))
      } else {
        .x <-
          mutate(.x, new_hisced = as.character(high_edu_broad))
      }
      .x
})

@@

For Denmark, for example, we learned a great deal about how to improve social mobility by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility for less advantaged children \citep{bjorklund2009, jaeger2007}. But the important finding came when we discovered the main reason behind their social escalator: its' educational system. For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impacts in helping children reach their fullest potential, specially disadvantaged children. Moreover, it tracks students into different curricula very late compared to other Europen countries Europe (age 16). These two traits make the Danish system uniquely effective. The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable. In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.

@@

\citet{reardon2011} finds that the increase in the gaps has occured predominantely from the 1970's until the 2000's. In fact, the hard numbers suggest that the gap widened by \~  40-50\%. The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of \~  75\%. Given that the studies before 1970 are less reliablle in terms of comparability and sampling design, the author computes all results for both before/after 1970. To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations in the year 2010. Using longitudinal data, \citet{bradbury2015} find similar results. Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25. Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of \~ 1.25-1.50 standard deviations. One important drawback of these studies is that they don't present the uncertainty of this estimates. Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy. It could be that the gap is at \~ 1.25 standard deviations, varying up to 1.75 and down to 0.80. Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.

@@

The American social landscape is very distinctive relative to Europe. The reasons the authors find a reversal in the trend could be numerous and should be studied very close. For example, the American educational system lacks any formal curricular differentiation \emph{\'a la} European style. It is true nonetheless that \citet{reardon2011} did not cover from the period 2000 onwards, which is specifically the trend they studied so we have no other comparison. \citet{reardon_portilla} suggests that the reversal is likely due to the high increase of preschool enrollment. They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially. This explanation still needs further confirmation.

Motivated by these recent results, other authors have taken this analysis to an international context in order to discover between-country trends. The work of \citet{bradbury2015} employs a unique comparative analysis of the achievement gap between Australia, United Kingdom, United States and Canada. Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is not only the first study that uses panel data to study achievement gaps, but to also do it between countries}. The core finding behind their study is that the American achievement gap is much wider than in any other comparison country, specially Australia and Canada. They find that once the achievement gap is present in early school entry, it doesn't seem to narrow or neither widen very much over the life course. In fact, they calculate that the schooling environment once they enter school can only explain about \~ 30-40\% of the high school SES gap. This suggests that once the achievement gap widens before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous RCT's show that high quality schooling can indeed help ease the gap, in some instances even close it \citep{campbell2002}.}. One exception is the United Kingdom, which they found to help close the gap in early primary years. This can likely be due to the comprehensive schooling and also the public support by the welfare state in dimensions like health and welfare support.

@@

More formally, We're interested in studying
a) Which countries are experiencing changes in the achievement gap;
b) the rate at which each country-gap is widening/narrowing and;
c) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
d) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;

@@

a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data. This shows that it is important to study the changes in the achievement gap over time. This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.

b) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year available. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}

c) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving. If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.

d) We want to test the notion that better performing countries have a smaller degree of inequality than other countries. \citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious, however. For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.

@@

This new standardized variable has a mean of zero. Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves because the test scores have now the same metric across time. However, if the variance of academic achievement changes over time, then standardizing the overall score at each country-wave pair actually makes the transformation biased. That is, by standardizing we're forcing the standard deviation of test scores to be zero across all waves. But if the true deviations of the median academic achievement changes over time, then the estimated trend in the SES gaps will be underestimated, or vice versa.

% Show graph of variance over time
% I'm showing standard deviations, is it better to show variance?

\begin{figure}
\begin{center}
<<variance_pisa, include = T, out.height = '3in', out.width = '3.5in', fig.align = 'center'>>=

@@

var_years %>%
  setNames(seq(2000, 2015, 3)) %>%
  enframer("name") %>%
  ggplot(aes(name, value)) + geom_col()
@
\caption{Standard deviation of test scores across all waves.}
\end{center}
\end{figure}

@@

Another concern is that if the gaps at different waves are measured with tests that have different amounts of measurement error, then the amount of bias will not be the same in each measure of the gap. This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citep{reardon2011}. PISA has tried to make sure the tests are comparable across waves \citep{pisa2012_technical}, but we still have to adjust for this imprecision.

@@

Finally, we calculate the SES achievement gap for each country by extracting the fitted \(\alpha\) and \(\beta\) for each country \(j\) and calculating the difference of the predicted score of high and low SES.

@@

<<escsdummy_charachteristics>>=
escs_dummy_creator <- function(df, probs) {

@@

escs_data[[1]]$high_edu_broad <- dplyr::recode(escs_data[[1]]$high_edu_broad, `6` = 7)

edu_calc <- function(df) {
  df %>%
  select(country, escs_dummy, high_edu_broad) %>%
  count(country, escs_dummy, high_edu_broad) %>%
  group_by(country, escs_dummy) %>%
  mutate(total_n = sum(n),
         perc = n / total_n * 100) %>%
  filter(country %in% countries, high_edu_broad %in% c(1, 7)) %>%
  select(country, escs_dummy, high_edu_broad, perc) %>%
  map_if(is_double, round, 2) %>%
  as_tibble() %>%
  filter(!is.na(escs_dummy)) %>%
  spread(high_edu_broad, perc) 
}

summary_data <-
  map(escs_data, edu_calc) %>%
  setNames(years) %>%
  enframer()
@

% Figure out how to ouput data frames as latex tables.
<<sample_size, results = 'tex'>>=
# Get sample counts for each dummy
sample_size_calc <- function(df, probs, selected = F, cnts = NULL) {

@@

sample_tables_topbottom <- sample_size_calc(adapted_year_data, c(.1, .9), selected = TRUE, countries)
sample_tables_topmid <- sample_size_calc(adapted_year_data, c(.5, .9), selected = TRUE, countries)
sample_tables_midbottom <- sample_size_calc(adapted_year_data, c(.1, .5), selected = TRUE, countries)

summary_table <-
  left_join(summary_data, sample_tables_topbottom) %>%
  filter(name %in% c("2000", "2015")) %>%
  arrange(country, name)

table_coming <-
  summary_table %>%
  select(-perc) %>%
  gather(cats, vals, -(name:escs_dummy)) %>%
  unite(all_vals, escs_dummy, cats, sep = "_") %>%
  spread(all_vals, vals) %>%
  map_at(c(3, 4, 6, 7), ~ {
    .x[is.na(.x)] <- 0
    paste0(round(.x, 0), "%")
  }) %>%
  map_if(is_double, as.character) %>%
  data.frame() %>%
  xtable::xtable()

addtorow <- list()
addtorow$pos <- list(0)
addtorow$command <- paste("\\hline \\
                          & & & Low SES & & & High SES \\\\
                          \\cmidrule(l){3-5}
                          \\cmidrule(l){6-8}",
                          paste0(
                            c("Year",
                              "Countries",
                              "\\% Low Edu",
                              "\\% High Edu",
                              "Sample size",
                              "\\% Low Edu",
                              "\\% High Edu",
                              "Sample size"),
                            collapse = " & "),
                          "\\\\ \\hline")

print(table_coming,
      add.to.row = addtorow,
      include.colnames = FALSE,
      include.rownames = FALSE,
      hline.after = NULL,
      caption.placement = "top",
      caption = "SES sample size and ISCED composition")
@

<<merge_math_read>>=

@@

% How to output data frames are latex tables?
<<rate_change>>=

@@

full_data %>%
  ggplot(aes(low_increase, high_increase), alpha = 0.2) +
  geom_rect(data = rect_data, aes(xmin = xst,
                                  xmax = xen,
                                  ymin = yst,
                                  ymax = yen),
            fill = rect_data$colour,
            alpha = 0.2,
            inherit.aes = FALSE) +
  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "grey", size = 1,
            linetype = "longdash") +
  geom_point(alpha = 0.2) +
  geom_point(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.7) +
  geom_line(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.4,
            stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "red", size = 1) +
  geom_text_repel(data = filter(full_data, continent == "my_cnt"),
                  aes(label = country), box.padding = unit(2.7, "lines")) +
  geom_vline(xintercept = 0, alpha = 0.5) +
  geom_hline(yintercept = 0, alpha = 0.5) +
  xlim(lims$xlim) +
  ylim(lims$ylim) +
  coord_cartesian(expand = FALSE) +
  annotate(geom = "text", x = 0.15, y = -0.2,
           label = "Low SES is catching up \n faster than High SES",
           fontface = 2, size = 3) +
  annotate(geom = "text", x = -0.05, y = 0.20,
           label = "High SES is increasing  \n faster than Low SES is catching up",
           fontface = 2, size = 3) +
  labs(x = "Average increase of low SES in SD", y = "Average increase of high SES in SD") +
  theme_minimal()
@

In a more convenient way, we can see that there is a lot of variance in the rate at which top/bottom groups are changing. We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as United States and Germany. We see a negative relationship where the faster the low SES catch up the slower the High SES decrease. Quite an interesting pattern because we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel). On the other hand, we do see more ocuntries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.

% How to output data frames are latex tables?
<<perc_increase_tables>>=
# Show the rates at which is increasing/decreasing
perc_increase_fun <- function(df) {
